---
title: "Project 1 - Powerlifting"
author: "Brandon Leslie"
format: 
  revealjs:
    theme: night
    scrollable: true
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, 
                      message = FALSE, 
                      warning = FALSE
                      )

```


```{r include = FALSE}
library(tidyverse)
library(tidymodels)
library(tinytex)
library(knitr)
power <- read_csv("/Users/blove/Downloads/DSC-260/Proj1/data/PowerRaw.csv")

```

```{r}

power <- power %>% 
  group_by(Name) %>% 
  mutate(
    BestBenchKg = max(Best3BenchKg),
    BestSquatKg = max(Best3SquatKg),
    BestDeadliftKg = max(Best3DeadliftKg),
    TotalKg = max(TotalKg)
  ) 
# Making new columns that show the best lift of all individuals. 

power <- power %>% 
  select(Name, Sex, Equipment, Age, BodyweightKg, BestSquatKg, BestBenchKg, BestDeadliftKg, TotalKg) %>%
  drop_na() %>%
  filter(Age >= 8) %>% 
  filter(TotalKg == max(TotalKg)) %>% 
  distinct(Name, .keep_all = TRUE)

power$Equipment <- as.factor(power$Equipment)

power$Equipment <- fct_relevel(power$Equipment, "Raw", "Wraps", "Single-ply", "Multi-ply", "Unlimited")

power <- power %>% 
  mutate(
    WeightClass = cut(BodyweightKg,
                      breaks = seq(0, 315, by = 45),
                      labels = paste(seq(0, 270, by = 45), seq(45, 315, by = 45), sep = "-"),
                      right = FALSE),
    AgeClass = cut(Age, 
                   breaks = c(0, 10, 15, 20, 30, 40, 50 , 60, 70, 80, Inf),
                   labels = c("0-9", "10-15", "16-20", "21-30", "31-40", 
                                      "41-50", "51-60", "61-70", "71-80", "81+"),
                           right = FALSE),
    p4p = (TotalKg / BodyweightKg)
)
# Creating new age/weight classes, also creating a new strength indicator (p4p)    
```


```{r}
q1_plot_weight <- function(data, x_var, y_var, title, x_label, y_label, maxY, lineY) {
  data %>%
    ggplot(aes_string(x = x_var, y = y_var)) +
    geom_bin_2d(aes(fill = Sex)) +
    geom_vline(xintercept = seq(0, 270, by = 45), color = "black", linewidth = .5, alpha = 0.8) +
    geom_hline(yintercept = seq(0, maxY, by = lineY), color = "black", linewidth = .5, alpha = 0.8) +
    geom_smooth(color = "white", method = "gam", se = FALSE, linetype = "dashed") +
    facet_wrap(~Equipment) +
    labs(
      title = title,
      subtitle = "Faceted by Equipment",
      x = x_label,
      y = y_label
    ) +
    theme_dark() +
    scale_x_continuous(breaks = breaks_width(45)) +
    scale_y_continuous(breaks = breaks_width(200), limits = c(0,maxY)) +
    scale_fill_manual(values = c("M" = "blue", "F" = "red", "Mx" = "green")) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1.5, linetype = "longdash"),
          axis.text = element_text(face = "bold"))
}

q1_plot_age <- function(data, x_var, y_var, title, x_label, y_label, maxY, lineY) {
  data %>%
    ggplot(aes_string(x = x_var, y = y_var)) +
    geom_bin_2d(aes(fill = Sex)) +
    geom_vline(xintercept = seq(0, 90, by = 10), color = "black", linewidth = .5, alpha = 0.8) +
    geom_hline(yintercept = seq(0, maxY, by = lineY), color = "black", linewidth = .5, alpha = 0.8) +
        geom_smooth(color = "white", method = "gam", se = FALSE, linetype = "dashed") + 
    facet_wrap(~Equipment) +
    labs(
      title = title,
      subtitle = "Faceted by Equipment",
      x = x_label,
      y = y_label
    ) +
    theme_dark() +
    scale_x_continuous(breaks = breaks_width(10)) +
    scale_y_continuous(breaks = breaks_width(200), limits = c(0,maxY)) +
    scale_fill_manual(values = c("M" = "blue", "F" = "red", "Mx" = "green")) +
    theme(panel.border = element_rect(color = "black", fill = NA, size = 1, linetype = "longdash"),
          axis.text = element_text(face = "bold")
) 
}

# Makes functions for the plots/leaders so code is more efficient. 
# Using slice_max instead of filter to keep the top 3 instead of just top 1.

weightclass_leaders <- function(data, lift_column, top_n = 3) {
  data %>%
    group_by(WeightClass, Sex) %>%
    slice_max(order_by = .data[[lift_column]], n = top_n, with_ties = FALSE) %>%
    arrange(desc(WeightClass), Sex, .data[[lift_column]]) %>%
    select(Name, all_of(lift_column), BodyweightKg, Sex, p4p, Age, Equipment) %>% 
        kable(caption = paste("Top", top_n, "Lifters by Weight Class"), digits = 2)

}

ageclass_leaders <- function(data, lift_column, top_n = 3) {
  data %>%
    group_by(AgeClass, Sex) %>%
    slice_max(order_by = .data[[lift_column]], n = top_n, with_ties = FALSE) %>%
    arrange((AgeClass), Sex, .data[[lift_column]]) %>%
    select(Name, all_of(lift_column), BodyweightKg, Sex, p4p, Age, Equipment) %>% 
        kable(caption = paste("Top", top_n, "Lifters by Age Class"), digits = 2)

}

leaders_general <- function(data, lift_column, top_n = 3) {
  data %>% 
  group_by(Sex, WeightClass) %>%
    slice_max(order_by = .data[[lift_column]], n = top_n, with_ties = FALSE) %>%
    arrange(desc(WeightClass), Sex, .data[[lift_column]]) %>%
    select(Name, all_of(lift_column), BodyweightKg, Sex, p4p, Age, Equipment) %>% 
        kable(caption = paste("Top", top_n, "Lifters Overall"), digits = 2)

}

```
# Introduction
## About The Data

<hr>

This report utilizes a database with ~3 million records of powerlifting meet data worldwide. Powerlifting, a globally and nationally competed sport, involves lifters maximizing weight in three key lifts over a set period (typically six months). These lifts, recognized as strength benchmarks, have existed for over a century.

## Key Lifts

These lifts include...

-   **Barbell Bench Press** \| An upper-body lift using the chest, triceps, and shoulders to press a barbell off the chest. A good lift is smooth, level, and free of excessive glute movement.

<hr>

-   **Barbell Back Squat**\| A lower-body lift engaging the quads, glutes, and core to squat a barbell from the shoulders. Proper form includes good posture, smooth motion, and depth (knee crease below 90°).

<hr>

-   **Barbell Deadlift**\| A full-body lift pulling a barbell from the ground to lockout. A good deadlift is smooth, with minimal jerking and proper posture.


## Questions

After inspecting this data, I have been left with two questions:

1.  **What is the difference between Age & Weight classes on their progression from Raw to Unlimited equipment lifts? Why might this be?**

<hr>

2.  **What is the difference of pound-for-pound strength versus total strength when looking at different weight classes? Why might this be?**

# Question One:
<hr>
**What is the difference between Age & Weight classes on their progression from Raw to Unlimited equipment lifts? Why might this be?**

## Overview

Aims to see how progressing equipment levels affect the performance of different Age & Weight classes.

## Q1 | Visulization - Weight 

Total Leaders/Visualizations by Weight  


```{r}
q1_plot_weight(power, "BodyweightKg", "TotalKg", "Total By Weight", "Weight-(kg)", "Total-(kg)", 1300, 200)

```

## Q1 | Data - Weight 

```{r}
weightclass_leaders(power, "TotalKg")
```


## Q1 | Visulization - Age


```{r}
q1_plot_age(power, "Age", "TotalKg", "Best Total by Age", "Age", "Total (kg)", 1300, 200)
```

## Q1 | Data - Age

```{r}
ageclass_leaders(power, "TotalKg")
```


## Conclusions

- Weight and age categories show substantial differences as lifters transition into better equipped lifts.

<hr>

- Equipment appears to have little effect on strength gains.

<hr>

- The greatest strength development is seen in the 10-20 years age group, with peaks in strength in the subsequent age group, then declining after that.

<hr>

- This is observed across all types of machinery.

<hr>

- In weight classes, Multi-ply and Unlimited equipment show more linear strength progress.

<hr>

- There is no clear advantage to equipping since most top players use very little equipment.
Overall, age and weight classes are more relevant to strength gains than equipment.


# Question Two:
<hr>
**What is the difference of pound-for-pound strength versus total strength when looking at different weight classes? Why might this be?**

## Overview
Aims to compare two different metrics of strength across different weight classes:

- **Total**
  - Cumulative weight of lifters lifts (Bench, Squat, Deadlift)

<hr> 

- **Pound-for-Pound Ratio** (p4p)
  - A lifters total divided by their body weight, great indicator of strength for those who may be genetically disadvantaged. 

## Q2 | Visulization - Total

```{r}
power %>% 
 ggplot(aes(x = BodyweightKg, y = TotalKg)) +
  geom_smooth(aes(color = Sex, linetype = Sex), method = "gam", se = FALSE) +
  geom_vline(xintercept = seq(0, 270, by = 45), color = "black", linewidth = .5, alpha = 0.8) +
  geom_hline(yintercept = seq(0, 1300, by = 200), color = "black", linewidth = .5, alpha = 0.8) +
  theme_dark() +
    scale_x_continuous(breaks = breaks_width(45)) +
    scale_y_continuous(breaks = breaks_width(200), limits = c(0,NA)) +
    scale_color_manual(values = c("M" = "blue", "F" = "red", "Mx" = "green")) +
  labs(
    title = "Total by Bodyweight",
    x = "Bodyweight-(kg)",
    y = "Total-(kg)"
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1.5, linetype = "longdash"),
        axis.text = element_text(face = "bold"))
```

## Q2 | Data - Total

```{r}
leaders_general(power, "TotalKg")
```


## Q2 | Visulization - p4p
```{r}
power %>% 
 ggplot(aes(x = BodyweightKg, y = p4p)) +
  geom_smooth(aes(color = Sex, linetype = Sex), method = "gam", se = FALSE) +
  geom_vline(xintercept = seq(0, 270, by = 45), color = "black", linewidth = .5, alpha = 0.8) +
  geom_hline(yintercept = seq(0, 8, by = 1), color = "black", linewidth = .5, alpha = 0.8) +
  theme_dark() +
    scale_x_continuous(breaks = breaks_width(45)) +
    scale_y_continuous(breaks = breaks_width(1), limits = c(0,8)) +
    scale_color_manual(values = c("M" = "blue", "F" = "red", "Mx" = "green")) +
  labs(
    title = "p4p Ratio by Bodyweight",
    x = "Bodyweight-(kg)",
    y = "p4p Ratio"
  ) +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 1.5, linetype = "longdash"),
        axis.text = element_text(face = "bold"))

```


## Q2 | Data - p4p
```{r}
leaders_general(power, "p4p")
```


## Conclusions

- Lower-weight classes tend to show greater p4p strength potential in elite lifters.

<hr>

- Higher-weight classes see a decrease in p4p strength but an increase in total strength.

<hr>

- p4p and total strength changes are not proportional but somewhat perpendicular.

<hr>

- Strength doesn’t directly correlate with weight but with the "type" of weight (e.g., muscular vs. adipose tissue).

<hr>

- A lifter with more active weight (muscle/skeleton) is likely stronger than one with passive weight (fat).

<hr>

- Weight is not a direct indicator of strength; too much weight can be detrimental.

<hr>

- In powerlifting meets where p4p is used, excessive weight can hinder performance, as it affects the p4p ratio rather than total lifted weight.

# [Data Sourced Here](https://www.kaggle.com/datasets/open-powerlifting/powerlifting-database)
